---
title: "test1"
author: "Syeda"
date: "5/27/2021"
output: html_document
---

```{r setup, eval=TRUE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r}
x = rnorm(10)
plot(x,pch=19,col="dodgerblue")
```

```{r}
y = rbinom(20,size=1,prob=0.5)
table(y)
```


Question 5
Load the Bottomly and the Bodymap data sets with the following code:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
pdata_bot=pData(bot)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)

Just considering the phenotype data what are some reasons that the Bottomly data set is likely a better experimental design than the Bodymap data? Imagine the question of interest in the Bottomly data is to compare strains and in the Bodymap data it is to compare tissues.



* The Bottomly data has biological replicates for each group but the Bodymap data does not. 


* The Bodymap data has measured more levels of the outcome of interest (tissues) than the Bottomly data has measured (strains).


* The Bottomly data set does not measure the age of the mice.


* Most of the tissues in the Bodymap data have a consistent number of technical replicates (2).


```{r 5}
library(Biobase)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
pdata_bot=pData(bot)

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata_bm=pData(bm)
```
Expected Answer6:
The Bottomly data has biological replicates for each group but the Bodymap data does not


Question 7
Load the data:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)

Which of the following code chunks will make a heatmap of the 500 most highly expressed genes (as defined by total count), without re-ordering due to clustering? Are the highly expressed samples next to each other in sample order?

```{r q7}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
```

```{r opt1}
row_sums = rowSums(edata)
edata = edata[order(-row_sums),]
index = 1:500
heatmap(edata[index,],Rowv=NA,Colv=NA)
```
Expected Ans7: 

row_sums = rowSums(edata)
edata = edata[order(-row_sums),]
index = 1:500
heatmap(edata[index,],Rowv=NA,Colv=NA)
yes they are


Question 8
Load the Bodymap data using the following code:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata = pData(bm)
edata = exprs(bm)

Make an MA-plot of the first sample versus the second sample using the log2 transform (hint: you may have to add 1 first) and the rlog transform from the DESeq2 package. How are the two MA-plots different? Which kind of genes appear most different in each plot?



* The plots look pretty similar, but the \verb|rlog|rlog transform seems to shrink the low abundance genes more. In both cases, the genes in the middle of the expression distribution show the biggest differences.


* The plots look pretty similar, but the \verb|log2|log2 plot seems to do a better job of shrinking the low abundance genes toward each other. In both cases, the genes in the middle of the expression distribution show the biggest differences.


* The plots look pretty similar, but there are two strong diagonal stripes (corresponding to the zero count genes) in the \verb|rlog|rlog plot. In both cases, the genes in the middle of the expression distribution show the biggest differences, but the low abundance genes seem to show smaller differences with the \verb|log2|log2 transform.


* The plots are very different, there are two strong diagonal stripes (corresponding to the zero count genes) in the \verb|log2|log2 plot and the high abundance genes are most different, but the low abundance genes seem to show smaller differences with the \verb|rlog|rlog transform.

```{r q8}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata = pData(bm)
edata = exprs(bm)
```

```{r 8-1,echo=FALSE}
# MA plot of the log2 transform of the first and second sample
mm = log2(edata[,1]+1)-log2(edata[,2]+1)
aa = log2(edata[,1]+1)+log2(edata[,2]+1)
plot(aa,mm,col=2)
```

```{r 8-2, echo=FALSE}
library(rlog)
# MA plot of the rlog transform of the first and second sample
rlog_edata = rlog(edata)
xx = rlog_edata[,1]+1 - rlog_edata[,2]+1
yy = rlog_edata[,1]+1 + rlog_edata[,2]+1

plot(yy,xx,col=3)
```

Expected Answer8:
The plots look pretty similar, but there are two strong diagonal stripes (corresponding to the zero count genes) in the log2 plot. In both cases, the genes in the middle of the expression distribution show the biggest differences, but the low abundance genes seem to show smaller differences with the rlog transform.


Question 9
Load the Montgomery and Pickrell eSet:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

Cluster the data in three ways:

1.With no changes to the data
2.After filtering all genes with rowMeans less than 100
3.After taking the log2 transform of the data without filtering

Color the samples by which study they came from (Hint: consider using the function myplclust.R in the package rafalib available from CRAN and looking at the argument lab.col.)

How do the methods compare in terms of how well they cluster the data by study? Why do you think that is?


Answer Choices:
* Clustering with or without log2 transform is about the same. Clustering after filtering shows better clustering with respect to the study variable. The reason is that it is just the lowly expressed genes that make the distance hard to calculate. 


* Clustering is identical with all three approaches and they show equal clustering. Clustering with or without log2 transform is about the same. Clustering after filtering shows better clustering with respect to the study variable. The reason is that it is just the lowly expressed genes that make the distance hard to calculate. 


* Clustering with or without log2 transform is about the same. Clustering after filtering shows better clustering with respect to the study variable. The reason is that the lowly expressed genes have some extreme outliers that skew the calculation. 


* Clustering with or without filtering is about the same. Clustering after the log2 transform shows better clustering with respect to the study variable. The likely reason is that the highly skewed distribution doesn't match the Euclidean distance metric being used in the clustering example.

```{r 9-1,echo=FALSE}

#Plot Settings
tropical = c('darkorange','dodgerblue','hotpink','limegreen','yellow')
palette(tropical)
par(pch=19)

#Load necessary packages
library(devtools)
library(Biobase)
library(dendextend)
library(rafalib)

#Load the data set
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

```

```{r cl1,echo=FALSE}
#Cluster with original data
dist1 <- dist(t(edata))

colramp = colorRampPalette(c(3,"white",2))(9)
heatmap(as.matrix(dist1),col=colramp,Colv=NA,Rowv=NA)

hclust1 <- hclust(dist1)
plot(hclust1,hang=-1)

col <- c(rep(2,60), rep(3,69))
myplclust(hclust1, hclust1$labels, lab.col = col, hang = 0.1)
```

```{r clust2, echo=FALSE}
#Cluster after filtering genes with rowMeans less than 100

#Step 1: Take filter the genes that have rowMeans less than 100 
edata2 <- edata[rowMeans(edata)>100,] 


#Step 2: Calculate the distance between samples/columnss (default does between rows , so we do the transpose of data set for samples/columns)
dist2 <- dist(t(edata2))
#Distance matrix of distance between every pair of samples in data set
dist2

#Step3: Heatmap
colramp = colorRampPalette(c(3,"white",2))(9)
heatmap(as.matrix(dist2),col=colramp,Colv=NA,Rowv=NA)

hclust2 <- hclust(dist2)
plot(hclust2)

#Plot labels at same level
plot(hclust2, hang=-1)

#Color the dendogram based on study
myplclust(hclust2, hclust2$labels, lab.col = col, hang = 0.1)
```

```{r clust3,echo=FALSE}
#Cluster after log2 transform without filtering
#Step 1: Take the log2 transform
edata3 <- log2(edata + 1) 

#Step 2: Calculate the distance between samples/columnss (default does between rows , so we do the transpose of data set for samples/columns)
dist3 <- dist(t(edata3))
#Distance matrix of distance between every pair of samples in data set
dist1

#Step3: Heatmap
colramp = colorRampPalette(c(3,"white",2))(9)
heatmap(as.matrix(dist3),col=colramp,Colv=NA,Rowv=NA)

hclust3 <- hclust(dist3)
plot(hclust3)

#Plot labels at same level
plot(hclust3, hang=-1)


#Color the dendogram based on study
myplclust(hclust3, hclust3$labels, lab.col = col, hang = 0.1)
```
Expected Answer9: Clustering with or without filtering is about the same. Clustering after the log2 transform shows better clustering with respect to the study variable. The likely reason is that the highly skewed distribution doesn't match the Euclidean distance metric being used in the clustering example.


Question 10
Load the Montgomery and Pickrell eSet:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

Cluster the samples using k-means clustering after applying the \verb|log2|log2 transform (be sure to add 1). Set a seed for reproducible results (use \verb|set.seed(1235)|set.seed(1235)). If you choose two clusters, do you get the same two clusters as you get if you use the \verb|cutree|cutree function to cluster the samples into two groups? Which cluster matches most closely to the study labels?

Answer Choices:

* They produce different answers. The k-means clustering matches study better. Hierarchical clustering would look better if we went farther down the tree but the top split doesn't perfectly describe the study variable. 


* They produce the same answers and match the study variable equally well. 


* They produce the same answers except for three samples that hierarchical clustering correctly assigns to the right study but k-means does not. 


* They produce different answers, with k-means clustering giving a much more unbalanced clustering. The hierarchical clustering matches study better. 


```{r 10, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

```

```{r km_clust1 }
#Take log transform of data
edata4 <- log2(edata + 1)

#Set seed
set.seed(1235)

# K-means clustering with 2 clusters
kmeans1 <- kmeans(edata4, centers = 2)

names(kmeans1)

#Look at cluster table
table(kmeans1$cluster)
```

```{r 10 km_clust2,echo = FALSE}
cutree(hclust1, k = 2)

```

Expected Answer 10:

They produce different answers. The k-means clustering matches study better. Hierarchical clustering would look better if we went farther down the tree but the top split doesn't perfectly describe the study variable. 