---
title: "W2Quiz"
author: "Syeda"
date: "7/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r parameters}
#Setup graphing parameters

tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")
palette(tropical)
par(pch=19)
```

```{r libraries,echo=FALSE}
library(devtools)
library(Biobase)
```

Question 1
Load the Montgomery and Pickrell eSet:

1234567
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
What percentage of variation is explained by the 1st principal component in the data set if you:

1. Do no transformations?
2. log2(data + 1) transform?
3. log2(data + 1) transform and subtract row means?

```{r load_data, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```


```{r no_transform, echo=FALSE}
# Percent Variance Explained with no transformation of data
svd1 <- svd(edata)
dim(edata)
dim(svd1$v)
dim(svd1$u)

plot(svd1$d,ylab="singlular value", col=2)
#Plot percent variance explained
plot(svd1$d^2/sum(svd1$d^2),ylab="Percent variance explained",col=2)
```
Variance Explained 1: 0.89


```{r transform1,echo=FALSE}
# Percent Variance Explained with log2(data + 1) transform
edata2 <- log2(edata + 1)
svd2 <- svd(edata2)

plot(svd2$d,ylab="singlular value", col=3)
plot(svd2$d^2/sum(svd2$d^2),ylab="Percent variance explained",col=3)

```
Variance Explained 2: 0.97

```{r transform2,echo=FALSE}
# Percent Variance Explained with log2(data + 1) transform - rowMeans
edata3 <- edata2 - rowMeans(edata2)
svd3 <- svd(edata3)

plot(svd3$d,ylab="singlular value", col=3)

```
Variance Explained 3: 0.35

Expected Answer1: a) 0.89 b) 0.97 c) 0.35



Question 2

Load the Montgomery and Pickrell eSet:
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

Perform the log2(data + 1) transform and subtract row means from the samples. Set the seed to 333 and use k-means to cluster the samples into two clusters. Use svd to calculate the singular vectors. What is the correlation between the first singular vector and the sample clustering indicator?

```{r libraries,echo=FALSE}
library(devtools)
library(Biobase)
library(dendextend)
```


```{r load_data,echp=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

```


```{r kmeans,echo=FALSE}
#Log to transform of edata
edata2 <- log2(edata + 1)
#Subtract rowmeans
edata3 <- edata2 - rowMeans(edata2)



#Use edata3 - data after log2 transform and subtracting rowMeans
#Cluster the samples using k-means clustering
set.seed(333)
#Since we want to cluster the samples into two clusters we must take transpose of edata3 when doing k-means clustering
km1 <- kmeans(t(edata3),centers=2)
#Calculate svd for edata3
svd3 <- svd(edata3)
#Compute the first singular vector
f_sing <- svd3$v[,1]
#Find the correlation between the first singular vector and the sample clustering indicator
cor(f_sing,km1$cluster)

```

Expected Answer2:0.87



Question 3
Load the Bodymap data with the following command


con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
Fit a linear model relating the first gene’s counts to the number of technical replicates, treating the number of replicates as a factor. Plot the data for this gene versus the covariate. Can you think of why this model might not fit well?

```{r load_libraries, echo=FALSE}
library(devtools)
library(Biobase)
library(broom)

```

```{r loadbm_data, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

```


```{r linear_regression,echo=FALSE}
#Convert edata into a matrix
edata <- as.matrix(edata)
#Fit linear model
lm1 <- lm(edata[1,] ~ pdata_bm$num.tech.reps)
lm1

#tidy(lm1)

#Plot the data of num replicates vs first gene counts
plot(pdata_bm$num.tech.reps,edata[1,], col =1)
abline(lm1, col=2)
```

Expected Answer3: There is only one data point with a value of 6 so it is likely that the estimated value for that number of technical replicates is highly variable.

and There are very few samples with more than 2 replicates so the estimates for those values will not be very good. 

Question 4
Load the Bodymap data with the following command

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
Fit a linear model relating he first gene’s counts to the age of the person and the sex of the samples. What is the value and interpretation of the coefficient for age?



```{r q4, echo=FALSE}

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

#Try this
lm3 <- lm(edata[1,]~pdata_bm$age + pdata_bm$gender)
# get the coefficients
lm3
```
Expected Answer4: -23.91. This coefficient means that there is an average decrease of 23.91 in the count variable per year within each gender. 

Question 5
Load the Montgomery and Pickrell eSet:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

Perform the log2(data + 1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the lm.fit function (hint: don't forget the intercept). What is the dimension of the residual matrix, the effects matrix and the coefficients matrix?

```{r q5, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

```

```{r q5,echo=FALSE}
#Fit many regressions at once (fit regression models for all genes at once)
#Do log2 transform of edata and convert to matrix
edata_m <- log2(as.matrix(edata + 1))

#Create model matrix
mod <- model.matrix(~ pdata$population)
fit <- lm.fit(mod,t(edata_m))
#Dimensions of residuals matrix
dim(fit$residuals)
#Dimensions of Effects matrix
dim(fit$effects)
#Dimensions of coefficients matrix
dim(fit$coefficients)

```

Expected Answer5: 
Residual matrix:  129 x 52580 

Effects matrix: 129 x 52580

Coefficients matrix: 2 x 52580 



Question 6
Load the Montgomery and Pickrell eSet:

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

Perform the log2(data + 1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the lm.fit function (hint: don't forget the intercept). What is the effects matrix?

```{r q6, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

```{r q6_1, echo=FALSE}
#Fit many regressions at once (fit regression models for all genes at once)
#Do log2 transform of edata and convert to matrix
edata_m <- log2(as.matrix(edata + 1))

#Create model matrix
mod <- model.matrix(~ pdata$population)
fit <- lm.fit(mod,t(edata_m))
head(fit$effects,2)
```

Expected answer6: The estimated fitted values for all samples for each gene, with the values for each gene stored in the columns of the matrix. 


Question 7
Load the Bodymap data with the following command

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

Fit many regression models to the expression data where age is the outcome variable using the lmFit function from the limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is the coefficient for age for the 1,000th gene? Make a plot of the data and fitted values for this gene. Does the model fit well?


```{r q7_1, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

```

```{r q7_2, echo=FALSE}
#Create model matrix
edata_bm <- as.matrix(edata)
mod2 <- model.matrix(~ pdata_bm$age)
```

```{r q7_3, echo=FALSE}
#transpose edata
edata_t <- t(edata)
edatat_df <- as.data.frame(edata_t)
#Filter to remove samples with missing values for age
edata_filt <- edatat_df %>% filter(pdata_bm$age != 0)
edata_filt_t <- t(edata_filt)
```


```{r q7_4, echo=FALSE}
library(limma)
fit_limma <- lmFit(edata_filt_t,mod2)
#Get coefficients for the 1000th gene
fit_limma$coefficients[1000,]

```

```{r q7_4, echo=FALSE}

mod <-model.matrix(~pdata_bm$age)
fit2 <- lm.fit(mod,t(edata_sub))

plot(edata)
plot(fit1$fitted.values[,1000])
```


```{r q7, echo=FALSE}
pdata_sub <- subset(pdata_bm, pdata_bm$age != "NA")
edata_sub <- edata[,row.names(pdata_sub)]
fit1 <- lmFit(edata_sub,mod2)
fit_vals <- fit1$coefficients[1000,]

#Get fitted values
mod <-model.matrix(~pdata_bm$age)
fit2 <- lm.fit(mod,t(edata_sub)

plot(edata)
plot(fit2$fitted.values[,1000])
```
Expected Answer7: 
-27.61. The model doesn't fit well since there are two large outlying values and the rest of the values are near zero.


Question 8
Load the Bodymap data with the following command
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

Fit many regression models to the expression data where age is the outcome variable and tissue.type|tissue.type is an adjustment variable using the lmFit function from the limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is wrong with this model?

```{r q8, echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

```

```{r q8_1, echo=FALSE}
#Fit adjusted model using lm.fit
mod_adj <- model.matrix(~pdata_bm$age + as.factor(pdata_bm$tissue.type))
fit_adj <- lm.fit(mod_adj,t(edata_sub))

```

```{r q8_2, echo=FALSE}
#Fit adjusted model using limma lmFit() and tissue.type as adjustment variable
mod_adj <- model.matrix(~pdata_bm$age + as.factor(pdata_bm$tissue.type))
fit_limma <- lmFit(edata_sub,mod_adj)
#Get the coefficients
fit_limma$coefficients[1,]
```
Expected Answer8: Since tissue.type is a factor variable with many levels, this model has more coefficients to estimate per gene (18) than data points per gene (16).

Question 9
Why is it difficult to distinguish the study effect from the population effect in the Montgomery Pickrell dataset from ReCount? 

```{r q_9,echo=FALSE}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)

#Look at pdata
head(pdata)
```

Expected Answer9: The effects are difficult to distinguish because each study only measured one population. 


Question 10
Load the Bodymap data with the following command

con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

Set the seed using the command set.seed(33353) then estimate a single surrogate variable using the sva function after log2(data + 1) transforming the expression data, removing rows with rowMeans less than 1, and treating age as the outcome (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is the correlation between the estimated surrogate for batch and age? Is the surrogate more highly correlated with race or gender?

```{r q10,echo=FALSE}
library(devtools)
library(Biobase)
library(sva)
library(bladderbatch)
library(snpStats)


con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)

```


```{r q10_1,echo=FALSE}
set.seed(33353)

#Log2 transform
edata_t <- log2(edata + 1)


# First remove rows that have missing values for age
pdata_sub2 <- subset(pdata_bm, pdata_bm$age != "NA")
edata_sub2 <- edata_t[,row.names(pdata_sub2)]

#Remove rows with rowMeans less than 1
edata_cen <- edata_sub2[rowMeans(edata_sub2)>1,]

mod = model.matrix(~as.factor(age), data=pdata_sub2)
fit1 <- lm.fit(mod,t(edata_cen))
```


```{r q10_2, echo=FALSE}
#Determine the batch variable 

#Build model matrix
mod1 <- model.matrix(~age, data = pdata_sub2)
#Build model to compare to
mod2 <- model.matrix(~1,data = pdata_sub2)

#Create covariates by running sva function(covariates created are potential batch effects)
sva1 <- sva(edata_cen,mod1,mod2,n.sv = 1)

#look at covariates
names(sva1)
dim(sva1$sv)


#Find correlation
cor(sva1$sv,pdata_sub2$age)

cor(sva1$sv,as.numeric(pdata_sub2$gender))

cor(sva1$sv,as.numeric(pdata_sub2$race))

```

Expected Answer10:Correlation with age: 0.20

More highly correlated with gender. 